---
title: "Preparing DEMs for soil mapping"
bibliography: references.bib
---

## Introduction

This document runs through some LiDAR DEM preparation steps that can be helpful when modelling soils and landscapes.

### Demo data

I am using a demonstration dataset that Pierre Roudier has kindly made available in his R package '[hillshader](https://github.com/pierreroudier/hillshader)'. The dataset `maungawhau_hr` is a 1 m resolution DEM of Maungawhau, in Auckland.

```{r}
library(terra)
library(hillshader)

mw <- rast(maungawhau_hr)
mw
```

```{r}
#| echo: false
#| column: screen
  
library(leaflet)
library(scico)

pal <- scico(30, palette = 'batlow')

leaflet() %>% 
  addTiles() %>% 
  addRasterImage(mw, opacity= 0.7, colors = pal)

# todo - add a couple of A-B profile line plots?
```

### On-disk storage

The next tool I will be using can't directly consume in-memory R objects like the one above, so lets write the DEM to a file it can access.

```{r}
writeRaster(mw, 'maungawhau_hr.tif', 
            datatype = 'FLT4S', gdal = ('COMPRESS=LZW'),
            overwrite = TRUE)
```

### Noise removal

Most high-resolution DEMs have a certain amount of local surface roughness. Some of that is random noise, some of it is related to vegetation patterns (like grassed areas) or landscape alterations (like tillage marks), and some is left over from standard production steps like building removal.

Much of this can be removed using the Feature Preserving Smoothing method of @lindsay2019, available in the [WhiteBoxTools](https://www.whiteboxgeo.com/) geospatial data analysis platform. This software can be used on the command line or accessed in programming environments using wrapper packages. One such package exists for R:

```{r}
library(whitebox)
```

The default settings for the Feature Preserving Smoothing tool have been shown to be effective in a broad range of situations, so let's apply them here:

```{r}
wbt_feature_preserving_smoothing(
  dem       = 'maungawhau_hr.tif',
  output    = 'maungawhau_hr_FPS.tif',
  filter    = 11,
  norm_diff = 15,
  num_iter  = 3,
  max_diff  = 0.5,
  zfactor = NULL,
  wd = getwd(),
  verbose_mode = FALSE,
  compress_rasters = TRUE,
  command_only = FALSE
)

mw_fps <- rast('maungawhau_hr_FPS.tif')
```

The effect of this tool is quite subtle; the best way to quickly check the output is visually, by generating before-and-after hillshades. This is done below using GDAL:

```{r}
source('gdal_env.R') # sets a few necessary environment variables

system2('gdaldem', args = c('hillshade', '-multidirectional',
                            'maungawhau_hr.tif',
                            'mw_hillshade.tif'))

system2('gdaldem', args = c('hillshade', '-multidirectional',
                            'maungawhau_hr_FPS.tif',
                            'mw_FPS_hillshade.tif'))

mw_hs <- rast('mw_hillshade.tif')
mw_fps_hs <- rast('mw_fps_hillshade.tif')
```

Zooming in a bit, the results clearly show a smoother surface post-processing, without the loss of larger scale structure that can occur when de-noising using low-pass filters.

```{r}
library(sf)
library(tmap)

palg <- scico(255, palette = 'grayC', direction = -1)

hss <- c(mw_hs, mw_fps_hs)
names(hss) <- c('Before', 'After')

tm_shape(hss, unit= 'm',
         bbox = st_bbox(c('xmin' = 2667800, 'xmax' = 2667950,
                          'ymin' = 6479100,  'ymax' = 6479250))) +
  tm_raster(palette = palg, style = 'cont', legend.show = FALSE) +
  tm_scale_bar(text.size = 1)
```

Another way to verify the effect is to look at slope values:

```{r}
mw_slp <- terrain(mw)
mw_fps_slp <- terrain(mw_fps)
slps <- c(mw_slp, mw_fps_slp)
names(slps) <- c('Before', 'After')

tm_shape(slps,
         bbox = st_bbox(c('xmin' = 2667800, 'xmax' = 2667950,
                          'ymin' = 6479100,  'ymax' = 6479250))) +
  tm_raster(palette = pal, style = 'fixed',
            breaks = c(0,3,7,11,15,25,35,42,60),
            labels = c('A', 'B', 'C-', 'C+' ,'D', 'E', 'F', 'G'),
            title = 'Slope Class',
            legend.is.portrait = FALSE,
            alpha = 0.8) + 
  tm_layout(legend.outside.position = 'bottom')
```

Post-processing, slope category patches have smoother edges and a more coherent structure.
